---
title: "Reproducible Research - Peer Assessment 1"
author: "develHector"
date: "04/18/2015"
output:
  html_document: default
  pdf_document: default
---
https://github.com/develHector/RepData_PeerAssessment1  

###Loading and preprocessing the data###
Download the file in case it's not on the current directory
```{r Download, echo=TRUE }
downloadFile <- "activity.zip";
url <- "https://d396qusza40orc.cloudfront.net/repdata/data/activity.zip";
if( !file.exists(downloadFile) ) download.file( url, downloadFile, method = "curl" );
```
Unzip it in memory, and process/transofrm the zip onto raw data
```{r Unzip, cache=TRUE, echo=TRUE }
dataFile <- "activity.csv";
FileCon <- unz( downloadFile, dataFile );
RawData <- read.csv( FileCon, header=TRUE, nrows=17568 );
```

###Mean total number of steps taken per day###
Calculate the total number of steps taken per day  (I'll remove intervals without data)
```{r Calculate0, echo=TRUE }
StepsPerDay <- with( RawData, tapply( steps, date, sum, na.rm = TRUE ) )  ; 
```

Make a histogram of the total number of steps taken each day
```{r Histogram, echo=TRUE, fig.height=3.8, fig.width=6.2, fig.align='center' }
library( lattice ) ; # I chose this just because of the parameterless nice colored graphs
histogram( StepsPerDay ) ;
```

Calculate and report the mean and median of the total number of steps taken per day  
```{r ReportMeanAndMedian, echo=TRUE }
mean( StepsPerDay )
median( StepsPerDay )
```

###Average daily activity pattern###

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken
```{r MeanStepsPerInterval, echo=TRUE, fig.height=3.8, fig.width=6.2, fig.align='center' }
# This time I'll use the base::aggregate method instead of tapply, just for fun
MeanStepsPerInterval <- aggregate( steps ~ interval, data=RawData, FUN=mean, na.rm = TRUE ) ;
library( lattice ) ; 
xyplot( steps ~ interval, data=MeanStepsPerInterval, type="l",
  ylab = "Avg num steps (across all days)",
  xlab  = "5 min interval" )
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r MeanStepsPerIntervalMax, echo=TRUE }
MaxSteps <- max( MeanStepsPerInterval$steps ) ;
Max <- MeanStepsPerInterval[ MeanStepsPerInterval$steps == MaxSteps,  ] ;
# There's no inquirying on the value, just the interval, that's why only that I'm reporting
print( Max$interval ) ; 
```


###Imputing missing values###
Calculate and report the total number of missing values in the dataset
```{r MissingValuesReport, echo=TRUE }
MissingValues <- RawData[ is.na( RawData$steps ),] ;
nrow( MissingValues ) ;
```

Devise a strategy for filling in all of the missing values in the dataset.

  * My strategy will be to fill the NA values with the Average mean of the interval from the whole complete dataset
```{r StrategyDevising, echo=TRUE }
# We aggregated the mean in previous essignment question, let's do the same for this requirement
MeanStepsPerInterval <- aggregate( steps ~ interval, data=RawData, FUN=mean, na.rm = TRUE ) ;
# Merge it on interval
CompleteData <- merge( RawData, MeanStepsPerInterval, by="interval" ) ;
# Replace NA values with the mean of that same interval, of course and being single steps, rounded 
CompleteData[ is.na( CompleteData$steps.x ), "steps.x" ] <-
    round( CompleteData[ is.na( CompleteData$steps.x ), "steps.y" ] ) ;
```

Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r CreateNew, echo=TRUE }
# Let's remove work columns and order it as the original one
CompleteData$steps <- CompleteData$steps.x;
CompleteData$steps.x <- NULL ;
CompleteData$steps.y <- NULL ;
CompleteData <- CompleteData[ order(CompleteData$date, CompleteData$interval), ] ;
```

Make a histogram of the total number of steps taken each day
```{r CompleteHistogram, echo=TRUE, fig.height=3.8, fig.width=6.2, fig.align='center' }
StepsPerDayComplete <- with( CompleteData, tapply( X=steps, INDEX=date, FUN=sum ) )  ; 
histogram( StepsPerDayComplete )
```

Calculate and report the mean and median total number of steps taken per day.
```{r CompleteMeanAndMedian, echo=TRUE }
mean(StepsPerDayComplete)
median(StepsPerDayComplete)
```

Do these values differ from the estimates from the first part of the assignment?

* Yes they do, as follows
```{r Comparison, echo=TRUE }
mean(StepsPerDayComplete) - mean(StepsPerDay)
median(StepsPerDayComplete) - median(StepsPerDay)
```

What is the NA impact on the estimates of the total daily number of steps?

* Due to my replacement strategy (based on just the specific 5-min intervals) my conclusions are:
1. NA replacement tend to increase the central limit of the study.
2. Total number of steps got increased.
3. Because of this, both mean and median moved up.
4. Also, histogram presents more of a normal distribution.
5. Median got closer to the mean, almost the same value.


#Activity diff between weekdays and weekends#

Create a new factor variable in the dataset indicating whether a given date is a weekday or weekend day.
```{r NewVariable, echo=TRUE }
# For if your language is not English as mine, adjust it with Sys.setlocale("LC_ALL", "English")
CompleteData$weekend =  factor( weekdays( as.Date( CompleteData$date ) ) %in% c("Saturday","Sunday"), 
                                  labels = c("weekend", "weekday") ) ;
```


Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r NewPlotPerWeekday, echo=TRUE, fig.height=4.8, fig.width=6.2, fig.align='center' }
library(lattice)
MeanStepsPerInterval <- aggregate( steps ~ interval + weekend, data=CompleteData, FUN=mean ) ;
xyplot( steps ~ interval | weekend, data=MeanStepsPerInterval, type="l", layout = c(1,2) )  ;
```

##Thanks guys!##


